---
title: "Sampling Distributions and Statistical Inference: Chapter 3 Exercises"
format: html
editor:
  markdown:
    wrap: 72
---

These solutions work through Exercises 3.1, 3.5(a), and 3.5(b) from
Chapter 3 (Gerber & Green, *Field Experiments*).

## Exercise 3.1: Important Concepts

### 3.1(a)

> **What is a standard error? What is the difference between a standard
> error and a standard deviation?**

A **standard error** is the standard deviation of a *sampling
distribution*. In the experimental context, it measures how much the
estimated ATE would vary from one hypothetical random assignment to the
next: it summarizes the spread of the estimates $\hat\theta_j$ across
all $J$ possible randomizations,

$$SE(\hat\theta) = \sqrt{\frac{1}{J}\sum_{j=1}^{J}\left(\hat\theta_j - \bar{\hat\theta}\right)^2}.$$

$$SE=\frac{SD}{\sqrt{N}}$$

A **standard deviation** describes the dispersion of a variable's values
around their mean — it can be computed for any set of numbers (e.g., the
potential outcomes of a group of subjects). For a variable $X$ with $N$
values and mean $\bar X$,

$$SD(X) = \sqrt{\frac{1}{N}\sum_{i=1}^{N}\left(X_i - \bar X\right)^2}.$$

The distinction is about *what* is varying. A standard deviation
describes variability in a set of data values; a standard error
describes variability in an *estimate* (a statistic) across repeated
samples or random assignments. A standard error is itself a standard
deviation, but specifically the standard deviation of an estimator's
sampling distribution. As such, the standard error typically shrinks as
the sample size grows, whereas the standard deviation of the underlying
outcomes does not.

### 3.1(b)

> **How is randomization inference used to test the sharp null
> hypothesis of no effect for any subject?**

The **sharp null hypothesis** states that the treatment effect is zero
for *every* subject: $Y_i(1) = Y_i(0)$ for all $i$. Its great analytic
convenience is that, if it is true, each unit's observed outcome equals
*both* of its potential outcomes. That means the single dataset we
observe constitutes the *complete* schedule of potential outcomes.

Randomization inference exploits this. Under the sharp null we can:

1.  Take the observed outcomes as fixed and known for every unit.
2.  Enumerate (or randomly sample) all possible treatment assignments
    that the experiment's randomization procedure could have produced.
3.  For each hypothetical assignment, recompute the test statistic
    (e.g., the estimated ATE), building up the sampling distribution of
    that statistic *under the null*.
4.  Compare the actually observed test statistic to this distribution.
    The $p$-value is the proportion of hypothetical assignments that
    yield a test statistic at least as extreme as the observed one (in
    one tail for a one-sided test, in both tails for a two-sided test).

Because it uses the exact set of possible randomizations, this method is
*exact*: it makes no appeal to large-sample or normality approximations
and applies to any sample size and any outcome type.

### 3.1(c)

> **What is a 95% confidence interval?**

A **95% confidence interval** is a range of values constructed by a
procedure that, over hypothetical replications of the experiment, would
bracket the true parameter (here, the ATE) 95% of the time. If we
imagine repeating the experiment under identical conditions and building
an interval each time, about 95 out of every 100 such intervals would
contain the true ATE.

The key interpretive point is that the *interval* varies from one
experiment to the next while the true ATE stays fixed. The 95% refers to
the long-run coverage rate of the *procedure*, not to a probability that
the true value lies inside any one realized interval.

### 3.1(d)

> **How does complete random assignment differ from block random
> assignment and cluster random assignment?**

All three are forms of random assignment, differing in the unit that is
randomized and in any structure imposed on the randomization:

| Design | How treatment is assigned |
|:-----------------------------------|:-----------------------------------|
| **Complete** | Exactly $m$ of the $N$ *individual* units are assigned to treatment, with every such allocation equally likely. No grouping. |
| **Block** | Units are first partitioned into blocks (strata) of similar units; complete random assignment is then carried out *within each block*, fixing the number treated per block. |
| **Cluster** | Whole pre-existing groups (clusters) are assigned as intact units; every member of a cluster shares the same treatment status. |

Block assignment guarantees balance on the blocking variable and, when
blocks group units with similar potential outcomes, tends to *reduce*
sampling variability relative to complete assignment. Cluster assignment
is often dictated by practical constraints and tends to *increase*
sampling variability when units within a cluster have similar potential
outcomes, because the effective number of independent units is the
number of clusters rather than the number of individuals.

### 3.1(e)

> **Experiments that assign the same number of subjects to the treatment
> group and control group are said to have a "balanced design." What are
> some desirable statistical properties of balanced designs?**

When the variances of the treated and untreated potential outcomes are
similar, splitting the sample evenly ($m \approx N/2$) tends to
**minimize the standard error** of the estimated ATE — allocating the
observations equally is approximately optimal for precision in that
case.

A balanced design also simplifies and improves **interval estimation**.
The method for constructing confidence intervals can produce intervals
that are too wide or too narrow when one group is much larger than the
other and the groups have unequal variances; equal group sizes eliminate
this complication, so the estimated confidence intervals behave well. In
short, balanced designs generally deliver good precision and
well-behaved uncertainty estimates without requiring advance knowledge
of which group has more variable outcomes.

## Exercise 3.5

Both parts use the schedule of potential outcomes from Table 2.1, now
imagining a design in which **one** village is assigned to treatment and
the remaining six serve as control.

```{r}
#| message: false
library(tidyverse)

tab21 <- tibble(
  village = 1:7,
  Y0 = c(10, 15, 20, 20, 10, 15, 15),
  Y1 = c(15, 15, 30, 15, 20, 15, 30)
)

tab21 |> mutate(tau = Y1 - Y0)
```

The true ATE is the average of the village-level effects:

```{r}
mean(tab21$Y1 - tab21$Y0)
```

### 3.5(a)

> **Using Table 2.1, imagine that your experiment allocates one village
> to treatment. Calculate the estimated difference-in-means for all
> seven possible randomizations.**

With $m = 1$ treated and $N - m = 6$ control, there are exactly
$\binom{7}{1} = 7$ possible randomizations. For the randomization that
treats village $i$, the estimated difference-in-means is the treated
village's $Y_i(1)$ minus the average $Y(0)$ of the other six villages:

$$\widehat{ATE}_i = Y_i(1) - \frac{1}{6}\sum_{j \neq i} Y_j(0).$$

```{r}
dim_one <- map_dbl(tab21$village, function(i) {
  tab21$Y1[i] - mean(tab21$Y0[tab21$village != i])
})

tibble(treated_village = tab21$village, est_ATE = round(dim_one, 3))
```

The seven estimates range from about $-0.83$ (treating village 1) to
about $15.83$ (treating village 3), a very wide spread relative to the
true effect of 5.

### 3.5(b)

> **Show that the average of these estimates is the true ATE.**

```{r}
mean(dim_one)
```

The seven difference-in-means estimates average to exactly **5**, which
equals the true ATE computed above. This confirms that the
difference-in-means estimator is **unbiased** under random assignment:
even though any single one-village randomization can be far from the
truth (from $-0.83$ up to $15.83$), the estimates are correct *on
average* across the randomization distribution.
