---
title: | 
    | Potential Outcomes and Randomised Experiments
author: "Ryan T. Moore"
date: 2026-07-21
date-format: iso
execute: 
  echo: true
format: 
  beamer:
    fonttheme: serif
    include-in-header:
      - text: | 
          \usepackage{wasysym} 
      - file: zzz_beamer_header.tex
    section-titles: true
    toc: true
institute:
  - American University
bibliography: "../admin/main.bib"
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| results: false
#| warning: false

library(gt)
library(here)
library(kableExtra)
library(tidyverse)
```


# Preliminaries 

## Preliminaries

\Large

- Accept GitHub invitation (click link in email)
- Upload `.txt` or `.R` or `.qmd` or  `.Rmd` of solutions  
(`.pdf`, but comments)
- How `.qmd`/`.Rmd` works (more to come?)

\pause 

>- PS1 for today
>- PS2 for Thursday
>- Work together, but every keystroke your own
>- Ask questions in the Slack

<!-- *** -->

<!-- ![Ramsey & Schafer, _The Statistical Sleuth_](figs/02-grid.pdf) -->


## Review Questions

\large

1. When can we observe both $Y_i(1)$ and $Y_i(0)$?  
2. What is this fact called?
3. When does the empirical difference-in-means estimator exactly equal the true, underlying ATE?
4. What are the two parts of SUTVA?

***

\LARGE

\begin{center}
Exercise in Potential Outcomes
\end{center}

# Statistical Refreshment

## Expectation

\large
$$E(Y) = \sum\limits_{i=1}^n [y_i \cdot p(y_i)]$$
\pause 

Calculate $E(\text{Ideology})$.

```{r}
#| echo: false
#| eval: true

t1 <- tibble(Respondent = 1:5,
             Ideology = c(3, -2, 3, 1, 1))

t1 |> gt() |> cols_align("center")
```

\pause 

$$1 \cdot \frac{2}{5} + 3 \cdot \frac{2}{5} + (-2) \cdot \frac{1}{5} = \frac{6}{5}$$ 


## Conditional Expectation

\large

$$E(Y | X = x) = \sum\limits_{i=1}^n [y_i \cdot p(y_i | X = x)]$$

\pause

Calculate $E(\text{Ideology} | \text{Dem} = 0)$.

```{r}
#| echo: false
#| eval: true

t2 <- tibble(t1, Dem = c(0, 1, 0, 0, 1))

t2 |> gt() |> cols_align("center")
```

\pause 

\pause 

$$1 \cdot \frac{1}{3} + 3 \cdot \frac{2}{3} = \frac{7}{3}$$  

<!-- | Respondent | Ideology | Dem | -->
<!-- | :--------: | :------: | :-: | -->
<!-- |     1      |    3     |  0  | -->
<!-- |     2      |    -2    |  1  | -->
<!-- |     3      |    3     |  0  | -->
<!-- |     4      |    1     |  0  | -->
<!-- |     5      |    1     |  1  | -->

## Conditional Expectation

\large

$$E(Y | X = x) = \sum\limits_{i=1}^n [y_i \cdot p(y_i | X = x)]$$

Calculate $E(\text{Ideology} | \text{Dem} = 0)$

|          |     | Dem?    |
| :------: | :-: | :-: | - |
|          |     | 0   | 1 |
|          | 3   | 2   | 0 |
| Ideology | 1   | 1   | 1 |
|          | -2  | 0   | 1 |

(tabled data, cells: counts of units with row/column scores)

## Conditional Expectation

\large

$$E(Y | X = x) = \sum\limits_{i=1}^n [y_i \cdot p(y_i | X = x)]$$

Calculate $E(\text{Ideology} | \text{Dem} = 0)$

|          |     | Dem?    |
| :------: | :-: | :-: | - |
|          |     | 0   | 1 |
|          | 3   | 0.4   | 0.0 |
| Ideology | 1   | 0.2   | 0.2 |
|          | -2  | 0.0   | 0.2 |

(tabled data, cells: probability unit has row/column scores)

\pause 

$$3 \cdot \frac{0.4}{0.6} + 1 \cdot \frac{0.2}{0.6} = \frac{8}{3}$$

## Unbiasedness

\Large

Does estimator return the truth, on average?

\pause 

$$E(\hat{\theta}) = \theta$$

\pause 
\vspace{5mm}

>- $E\left[ \widehat{\overline{Y_1 - Y_0}} \right] = \overline{Y_1 - Y_0}$
>- $E\left[ \widehat{\beta_1} \right] = \beta_1$

<!-- E.g., Avg est of ATE $=$ true ATE -->

## Unbiasedness

\Large

What estimator used to estimate $\overline{Y_1 - Y_0}$?

\pause 
\vspace{5mm}

The difference-in-means estimator.

\pause 
\vspace{5mm}

$$\widehat{\overline{Y_1 - Y_0}} = \left(Y_1 | T_i = 1 \right) - 
\left(Y_0 | T_i = 0 \right)$$

\pause 
Is it a "good" estimator?

<!-- Strictly, these should be t_i = 1, e.g. (see GG) -->

## Unbiasedness

\Large 

The difference-in-means estimator is _unbiased_ for the true average treatment effect. 

\pause 
\vspace{10mm}

(See `02-unbiased.R`)


## Probability

\large

The Three Axioms

1. $P(A) \geq 0$
2. $P(\Omega) = 1$
3. If events _mutually exclusive_ (or, sets _disjoint_), then $$P(A \text{ or } B) = P(A) + P(B)$$


## Conditional Probability

\large

For events $A$ and $B$, 

\begin{eqnarray*}
P(A \text{ and } B) & = & P(A) P(B|A)\\
& = & P(B) P(A|B)
\end{eqnarray*}

Divide both sides by marginal probability $P(B)$ yields

\begin{eqnarray*}
P(A|B) & = & \frac{P(A \text{ and } B)}{P(B)}
\end{eqnarray*}

## Statistical Independence

\large

$A$ and $B$ are independent iff both

$$P(A | B) = P(A)$$ and $$P(B|A) = P(B)$$

## Conditional Independence

\large

$A$ and $B$ can be independent _conditional on $C$_.  

\pause 
\vspace{4mm}

Many times, _only_ independent given $C$.

\pause 
\vspace{4mm}

E.g., $Y_1, Y_0$ indep of $T$, given $X$.

\pause 
\vspace{4mm}

$$P(A \text{ and } B |C) = P(A|C)P(B|C)$$

## Potential Outcomes Model: Estimands

- Individual TE
\begin{displaymath}
\tau_i = Y_i(1) - Y_i(0)
\end{displaymath}

- Average treatment effect
	\begin{displaymath}
	ATE = E(Y_1 - Y_0) = E(Y_1) - E(Y_0)	
	\end{displaymath}

- Average treatment effect for the treated
	\begin{displaymath}
	ATT = E(Y_1 | T = 1) - E(Y_0 | T = 1)
	\end{displaymath} 


## Potential Outcomes Model: Estimands

The average treatment effect (ATE):

\begin{eqnarray*}
E(Y_{1} - Y_{0}) & =&  \frac{1}{2n} \sum\limits_{i=1}^{2n} \left( Y_{i1} -
  Y_{i0} \right) \\
  & = & \frac{1}{2n} \sum\limits_{i=1}^{2n} \left( Y_{i1} \right) -
  \frac{1}{2n} \sum\limits_{i=1}^{2n} \left(Y_{i0} \right)\\
& = & E(Y_{1}) - E(Y_{0}) 	%\\
%& = & \frac{E(Y_1 | T_i =1 ) + E(Y_1 | T_i =0 )}{2} - \\
%&& \quad \frac{E(Y_0 | T_i = 1) + E(Y_0 | T_i = 0)}{2}
\end{eqnarray*} 



## Potential Outcomes Model: Estimands

- If we know $(Y_1, Y_0)$ indep of $T$ 
<!-- (as random assignment of Tr promotes),  \pause -->

- Then,

\begin{eqnarray*}
E\left(Y_{1} \right) &=& E(Y_{1} | T = 1)\\ E\left(Y_{0} \right) & = & E(Y_{0} | T = 0)	
\end{eqnarray*}

- Then, can substitute 

\begin{eqnarray*}
	ATE & = & E(Y_{1}) - E(Y_{0}) \\
	& = & E(Y_1 | T = 1) - E(Y_0 | T = 0)
\end{eqnarray*}

\pause 

Observed diff in Tr and Co group means gives ATE!


## Potential Outcomes Model: Estimands

\large

@holland86: "_prima facie effect_": $$E(Y_t | S = t) - E(Y_c | S = c)$$

\pause 

"It is important to recognize that $E(Y_t)$ and $E(Y_t | S=t)$ are _not_ the same thing \ldots"


## Potential Outcomes Model: Estimands, Interpretation

@gergre12:

When $Y(1)$ and $Y(0)$ indep of $T$, 
\begin{eqnarray*}
ATE & = & E(Y_i(1) | T_i = 1) - E(Y_i(0) | T_i = 0) \\ \pause 
& = & E(Y_i(1)| T_i = 1) \alert{- E(Y_i(0) | T_i = 1) +} \\
&& \quad \alert{E(Y_i(0) | T_i = 1)} - E(Y_i(0) | T_i = 0) \\ \pause 
& = & \left[E(Y_i(1)| T_i = 1) - E(Y_i(0) | T_i = 1) \right] +\\
&& \quad \left[E(Y_i(0) | T_i = 1) - E(Y_i(0) | T_i = 0) \right] \\ \pause 
& = & E(Y_i(1) - Y_i(0) | T_i = 1) + \\
&& \quad E(Y_i(0) | T_i = 1) - E(Y_i(0) | T_i = 0) \\
\end{eqnarray*}

\pause 

$$\underbrace{E(Y_i(1) - Y_i(0) | T_i = 1)}_{\text{ATT}} + \underbrace{E(Y_i(0) | T_i = 1) - E(Y_i(0) | T_i = 0)}_{\text{Selection Bias}} $$

# Assignment of Treatment

## What can be a Treatment?

\Large

<!-- >- Causal effects: relative to some other condition -->
>- Timing clearly defined
>- Covariates: causally prior to treatment
>- "Attributes", "immutable characteristics": difficult to isolate
<!-- >- Dose-response "biological gradient" evidence -->

\pause 

@senwas16 : "Race as a Bundle of Sticks: Designs that Estimate Effects of Seemingly
Immutable Characteristics" (_elements_ of attributes varyingly manipulable)


## Attributes

"Causal effect of race"?

```{r}
#| eval: true
#| label: readresume
#| message: false
#| warning: false

data(resume, package = "qss")
dim(resume)
```

\pause 

```{r}
#| eval: true
#| warning: false

kable(table(resume$race, resume$call))
```

\pause

@bermul04: "Are Emily and Greg More Employable Than Lakisha and Jamal? A Field Experiment on Labor Market Discrimination"



## Random Assignment Mechanisms

\Large

>- Simple/complete randomisation  
$\quad$ (Bernoulli trial, prob $\pi$)
>- Complete randomisation / random allocation  
$\quad$ (fixed proportion to tr)
>- Blocked randomisations  
$\quad$ (fixed proportion to tr, w/in group)
>- Cluster randomisations  
$\quad$ (assignment at higher level)
>- Waitlists, stepped wedges, etc.

## The Potential Outcomes Model: Assignment

\large

Observed outcome: $Y_i = Y_i(1)\cdot T_i + Y_i(0)\cdot (1 - T_i)$ 

\vspace{7mm}

The assignment mechanism _selects_ which potential outcome we observe.

\pause 

\vspace{5mm}

\normalsize

(@gergre12 use $Y_i = Y_i(1)\cdot d_i + Y_i(0)\cdot (1 - d_i)$ to highlight that we observe pot outcome from treatment **actually** taken, not hypothetical or assigned treatment.)

## The Potential Outcomes Model: Assignment

\Large

\begin{center}
``Assignment mechanisms" are really missing-data-generating procedures.
\end{center}

## Ignorability

Assignment mechanism is _ignorable_ if $Y_{mis}$ conditnly indep of $T$

$$P(T | X, Y_{obs}, Y_{mis}) = P(T | X, Y_{obs})$$

\pause 

\vspace{5mm}

Nothing in unobserved $Y_{mis}$ informs relationship betw'n $Y_{obs}$, $T$.

## Unconfoundedness

\large

Some ignorable mechanisms are _unconfounded_, too.

$$P(T | X, Y_{obs}, Y_{mis}) = P(T| X)$$

\pause 

\vspace{5mm}

Nothing in $Y$ informs $T$.

\pause 

\vspace{5mm}

These are special (key!) cases of conditional independence.

## An Assignment Mechanism

Little & Rubin (2000):

| Patient | Y   | T   |
| :-----: | :-: | :-: |
| 1       | 6   | 1   |
| 2       | 12  | 1   |
| 3       | 9   | 0   |
| 4       | 11  | 0   |

\pause 

Clearly, treatment is harmful.  $\overline{Y(1)} - \overline{Y(0)} = 9 - 10 = -1$.

## An Assignment Mechanism

Little & Rubin (2000):


| Patient | Y(0)        | Y(1)         | $\tau$       | T |
| :-----: | :---------: | :----------: | :----------: | - |
| 1       |  | 6            |   | 1 |
| 2       |  | 12           |   | 1 |
| 3       | 9           |   |  | 0 |
| 4       | 11          |  |  | 0 |
||||||
| Mean    | 10          | 9            |   |   |


Clearly, treatment is harmful.  $$\overline{Y(1) | T = 1} - \overline{Y(0) | T = 0} = 9 - 10 = -1$$


## An Assignment Mechanism: Perfect Doctor

Little & Rubin (2000):

| Patient | Y(0)        | Y(1)         | $\tau$       | T |
| :-----: | :---------: | :----------: | :----------: | - |
| 1       | \alert{(1)} | 6            |   | 1 |
| 2       | \alert{(3)} | 12           |   | 1 |
| 3       | 9           | \alert{(8)}  |  | 0 |
| 4       | 11          | \alert{(10)} |  | 0 |
||||||
| Mean    | 10          | 9            |   |   |


\invisible{

Clearly, treatment is beneficial:  $$\overline{Y(1)} - \overline{Y(0)} =9  - 6 = 3$$
}


## An Assignment Mechanism: Perfect Doctor

Little & Rubin (2000):


| Patient | Y(0)        | Y(1)         | $\tau$       | T |
| :-----: | :---------: | :----------: | :----------: | - |
| 1       | \alert{(1)} | 6            | \alert{(5)}  | 1 |
| 2       | \alert{(3)} | 12           | \alert{(9)}  | 1 |
| 3       | 9           | \alert{(8)}  | \alert{(-1)} | 0 |
| 4       | 11          | \alert{(10)} | \alert{(-1)} | 0 |
||||||
| Mean    | 10          | 9            | \alert{(3)}  |   |


\pause 

Clearly, treatment is beneficial:  $$\overline{Y(1)} - \overline{Y(0)} =9  - 6 = 3$$

\pause 

This assg mechanism is non-ignorable, confounded.


# Randomisation (Design-based) Inference

***

\Huge

\begin{center}
A volunteer?
\end{center}

\pause 

\large

\vspace{3mm}

The task: select the 2 folders with messages \pause

\vspace{3mm}
>- What is our baseline expectation/model for this process?  
>    - ``No x-ray vision. No ESP. Effect of messages on choice $=0$.\text{''}
>- What is an alternative?
>    - ``Some way to detect messages. Message location $\to$ choice.\text{''}

\pause 

\vspace{3mm}

Select!

## Randomisation Inference

The possible choices:
\begin{center}
$\blacksquare \blacksquare \square \square \square$ \quad $\square \blacksquare \square \blacksquare \square$\\
$\blacksquare \square \blacksquare \square \square$ \quad $\square \blacksquare \square \square \blacksquare$\\
$\blacksquare \square \square \blacksquare \square$ \quad \alert<2->{$\square \square \blacksquare \blacksquare \square$}\\
$\blacksquare \square \square \square \blacksquare$ \quad $\square \square \blacksquare \square \blacksquare$\\
$\square \blacksquare \blacksquare \square \square$ \quad $\square \square \square \blacksquare \blacksquare$\\
\end{center}

\pause 

\begin{itemize}
	\item You chose $\_\_\_$ and $\_\_\_$.  Let $X=$ number found. \pause 
	\item What was $P(X \geq 2 | \text{no ESP})$?  \pause $\frac{1}{10} = 0.1$ \pause 
	\item What was $P(X \geq 1 | \text{no ESP})$?  \pause $\frac{7}{10} = 0.7$ \pause 
	\item What is ``prob result at least this extreme, given model of no effect\text{''}?  \pause 
	\item Definition of $p$-value! \pause 
	\item Valid, exact, with no distributional assumption, no large $n$.  \pause
	\item {\it Randomisation} creates dist'n of possible numbers correct 
\end{itemize}

## The Randomisation Distribution of $X$

\begin{center}
\only<1>{\includegraphics[angle=0, width=4in]{figs/02-riHistogram}}
\only<2>{\includegraphics[angle=0, width=4in]{figs/02-riHistogram2}}
\only<3>{\includegraphics[angle=0, width=4in]{figs/02-riHistogram3}}
\end{center}


## Parametric Null Hypothesis Significance Testing

\Large

- Specify and assume $H_0$
- Define $H_A$
- Examine reference dist'n ($t$, $\chi^2$, \ldots) under $H_0$
- Calculate $p$-value
- Compare to some $\alpha$; reject $H_0$ if $p<\alpha$



## Randomisation Inference

\Large 

>- Specify and assume $H_0$  
$\quad$ (sharp null of no treatment effect)
>- Define $H_A$
>- Create reference dist'n from all possible values of $X$ under $H_0$  
$\quad$ (or at least a big sample of them)
>- What prop. of possible values are ``at least as extreme as\text{''} observed?  
$\quad \rightsquigarrow$ $p$-value!
>- Compare to some $\alpha$; reject $H_0$ if $p<\alpha$
>- CA ballot ordering effects (JASA 2006)


## Randomisation Inference

\large 

\center
The RI $p$-value is $$p = \frac{\text{\# outcomes } \geq \text{as extreme as obs}}{\text{total \# outcomes}}$$

\pause 

or

$$p = \frac{\text{\# randomisations producing extreme } \widehat{ATE}}{\text{total \# randomisations}}$$

\pause 

\vspace{2mm}

How many randomisations are there?


## Combinations: Counting selected sets

How many ways to **select** $k$ things from a set of $n$ things?

$$_n C_k = \binom{n}{k} = \frac{_n P_k}{k!} = \frac{n!}{k!(n-k)!}$$
\pause

Suppose 5 units, $A$, $B$, $C$, $D$, $E$.

>- How many ways to order? $_n P_k$: $ABCDE$, $ABCED$, \ldots $= 5 \cdot 4 \cdot 3 \cdot 2 \cdot 1 = 5! = 120$
>- If only want 3 of 5? Divide by $2! = (n-k)!$  
(removing permutations from last 2 slots)
>- What if order doesn't matter? Divide by $3! = k!$  
(6 permutations for $ABC$, but only one combination)



## Combinations: Counting selected sets

How many ways to choose 5 villages of 10 for treatment? \pause 

$$_{10} C_5 = \binom{10}{5} = \frac{10!}{5!(10-5)!}$$
\pause 

$$\frac{10\cdot 9 \cdot 8 \cdot 7 \cdot 6}{5\cdot 4 \cdot 3 \cdot 2 \cdot 1} = 252$$


## Common Assumptions, Null Hypotheses

\large

\begin{itemize}
\item Constant effect:

$$\tau_i = Y_{i1}-Y_{i0} = \tau \quad \forall i$$

\item Null hypothesis of no average effect:

$$ATE = \overline{\tau} = 0$$

\item Sharp null hypothesis of no effect: 

$$\tau_i = 0$$
\end{itemize}


## An Assignment Mechanism: Perfect Doctor

\large 

Calculate RI $p$-value for Perfect Doctor, under sharp null. 


| Patient | Y(0)        | Y(1)         | $\tau$       | T |
| :-----: | :---------: | :----------: | :----------: | - |
| 1       | \alert{(1)} | 6            | \alert{(5)}  | 1 |
| 2       | \alert{(3)} | 12           | \alert{(9)}  | 1 |
| 3       | 9           | \alert{(8)}  | \alert{(-1)} | 0 |
| 4       | 11          | \alert{(10)} | \alert{(-1)} | 0 |
||||||
| Mean    | 10          | 9            | \alert{(3)}  |   |


\pause 
\vspace{3mm}

(See `02-ri-perfect-dr.R`)

## RI versus the $t$-test

\Large

Perfect Doctor:

- RI: $p = 1$
- `t.test()`: $p \approx 0.8$
- "If no tr effect, then this result typical"

\pause

\vspace{4mm}

(Odd logic of NHST: "assume false thing, how strange is data?")


## Randomisation Inference, Example 2

Assumed table of potential outcomes:

\begin{center}
\begin{tabular}{ccccc}
	Village & $T$ & \% if Female & \% if Male &  $\tau_i$ \\
	& &  Head, $Y(1)$ &  Head, $Y(0)$ &  \\ \hline
	1 & $\_$ & 15 & 10 & 5\\
	2 & $\_$ & 15 & 15 & 0 \\
	3 & $\_$ & 30 & 20 & 10 \\
	4 & $\_$ & 15 & 20 & -5\\
	5 & $\_$ & 20 & 10 & 10 \\
	6 & $\_$ & 15 & 15 & 0 \\
	7 & $\_$ & 30 & 15 & 15\\ \hline
Average	 && 20 & 15 & 5
\end{tabular}
\end{center}


## Randomisation Inference, Example 2

Suppose we randomly select 2 villages to have female-headed councils, and observe

\begin{center}
\begin{tabular}{ccccc}
	Village & $T$ & \% if Female & \% if Male &  $\tau_i$ \\
	& &  Head, $Y(1)$ &  Head, $Y(0)$ &  \\ \hline
	1 & F & 15 & & \\
	2 & M &  & 15 &  \\
	3 & M &  & 20 &  \\
	4 & M &  & 20 & \\
	5 & M &  & 10 &  \\
	6 & M &  & 15 & \\
	7 & F & 30 &  & \\ \hline
Average	 && 22.5 & 16 & 6.5
\end{tabular}
\end{center}


## Randomisation Inference, Example 2

We assume the _sharp null_ hypothesis (assumed values in \textcolor{red}{red}): 

\begin{center}
\begin{tabular}{ccccc}
	Village & $T$ & \% if Female & \% if Male &  $\tau_i$ \\
	& &  Head, $Y(1)$ &  Head, $Y(0)$ &  \\ \hline
	1 & F & 15 & {\color{red} 15} & {\color{red} 0}\\
	2 & M & {\color{red} 15} & 15 & {\color{red} 0} \\
	3 & M & {\color{red} 20} & 20 & {\color{red} 0} \\
	4 & M & {\color{red} 20} & 20 & {\color{red} 0}\\
	5 & M & {\color{red} 10} & 10 & {\color{red} 0} \\
	6 & M & {\color{red} 15} & 15 &{\color{red} 0} \\
	7 & F & 30 & {\color{red} 30} & {\color{red} 0} \\ \hline
Average	 &&  &  & {\color{red} 0}
\end{tabular}
\end{center}


## Randomisation Inference, Example 2

Then we estimate what the observed ATE would be for all the possible random assignments.

First,

\begin{center}
\begin{tabular}{ccccc}
	Village & $T$ & \% if Female & \% if Male &  $\tau_i$ \\
	& &  Head, $Y(1)$ &  Head, $Y(0)$ &  \\ \hline
	1 & F & 15 & & \\
	2 & M &  & 15 &  \\
	3 & M &  & 20 &  \\
	4 & M &  & 20 & \\
	5 & M &  & 10 &  \\
	6 & F & {\color{red} 15} &  & \\
	7 & M &  & {\color{red} 30} &  \\ \hline
Average	 && 15 & 19 & -4
\end{tabular}
\end{center}


## Randomisation Inference, Example 2

Second,

\begin{center}
\begin{tabular}{ccccc}
	Village & $T$ & \% if Female & \% if Male &  $\tau_i$ \\
	& &  Head, $Y(1)$ &  Head, $Y(0)$ &  \\ \hline
	1 & F & 15 &  & \\
	2 & M &  & 15 &  \\
	3 & M & & 20 &  \\
	4 & M &  & 20 & \\
	5 & F & {\color{red} 10} &  &  \\
	6 & M &  & 15 & \\
	7 & M & & {\color{red} 30} &  \\ \hline
Average	 && 12.5 & 20 & -7.5
\end{tabular}
\end{center}


## Randomisation Inference, Example 2

\ldots, \pause and all the others.  The full set of $\frac{7!}{2!5!} = 21$ differences in means:

\begin{center}
\begin{tabular}{ccc}
& Estimate & Frequency  \\ \hline
& -7.5 & 3 	   \\
& -4 & 5 \\
& -0.5 & 6 \\
& 3 & 2 \\
& 6.5 & 3 \\
& 10 & 2 \\ \hline
Total & & 21
\end{tabular}
\end{center}

\pause

How many are at least as extreme as my 6.5? \pause

\vspace{2mm}
Two-sided (``women's \% $\neq$ men's\text{''}): $p = \frac{8}{21} \approx 0.38$ \pause

\vspace{2mm}
One-sided (``women's \% $>$ men's\text{''}): $p = \frac{5}{21} \approx 0.24$


## Randomisation Inference

\Large

These $p$-values

>- are exact, not approximate
>- do not require asymptotic large $N \to \infty$
>- need no distrib assump (normal as $N \to \infty$)
>- do not require calculating standard error (SE)
>- _Randomisation_ creates distribution of possible numbers correct/treatment effects
>- can be approximated by sampling
>    - Assigning 500 of 1000 respondents to treatment?  Randomly generate 10,000 assignments from $2.7 \times 10^{299}$ possible \ldots
>- can be computationally intensive


## Randomisation Inference

\large

- Resume audit study, @bermul04

```{r}
#| echo: false
#| warning: false

kable(table(resume$race, resume$call))
```

\pause 

- Only possible values: $\tau_i \in \{-1, 0, 1\}$

\pause 

```{r}
#| message: false
#| warning: false

resume |> group_by(race) |>
  summarise(call_rate = mean(call))
```


***

\large

>- Assume the sharp null $\tau_i = 0$ for every employer.
>- $H_0: \mu_{\text{black name}} = \mu_{\text{white name}}$
>- $H_A: \mu_{\text{black name}} \neq \mu_{\text{white name}}$
>- Create reference dist'n of all possible assignments

\pause 

$$_{4870} C_{2435} = \binom{4870}{2435} = \frac{4870\cdot 4869 \cdot \ldots \cdot 2436}{2435!}$$ 

\pause 

$\approx 1.1 \times 10^{1464}$

\pause 

(There are $\approx 10^{86}$ fundamental particles in the universe.)

\pause 

>- Let's do 1000, or 100,000 -- something reasonable
>- See `02-ri-resume-donate.R`


***

\begin{center}
\only<1>{\includegraphics[angle=0, width=3.5in]{figs/02-riHist_resume1}}
\only<2>{\includegraphics[angle=0, width=3.5in]{figs/02-riHist_resume2}}
\end{center}

## Randomisation Inference

\Large

- Gerber & Green donations example, p. 65
- Possible values $\tau_i \in (-\infty, \infty)$
- $Y_1$, $Y_0$, $\tau$ likely very skewed
- See `02-ri-resume-donate.R`


***

\begin{center}
\only<1>{\includegraphics[angle=0, width=3.5in]{figs/02-riHist_donate1}}
\only<2>{\includegraphics[angle=0, width=3.5in]{figs/02-riHist_donate2}}
\end{center}

## The RI Confidence Interval

\large

Recall that 

\begin{center}
``reject $H_0$ at $\alpha = 0.05$'' $\equiv$ ``$H_0$ falls outside 95\% CI''
\end{center}

\pause 
\vspace{5mm}

Create RI confidence ints (``invert the randomisation test\text{''}):

>- Posit $H_0: \tau = \tau^* \in \{ \ldots, -2, -1, 0, 1, 2, \ldots \}$
>- RI test whether to reject $H_0$
>- If not, then $\tau^*$ is in CI
>- CI consists of set of $\tau^*$ not unusual, given data
>- See `02-ri-resume-donate.R`


***

\huge

\begin{center}
Next:\\
Covariates in Experiments\\
PS2 due 
\end{center}



## References {.allowframebreaks}

\footnotesize