---
title: "Problem Set 2"
author: "Wannes De Meyer"
date: "2026-07-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 3.5 (Gerber & Green, 2012)

#### Using Table 2.1, imagine that your experiment allocates one village to treatment.

Packages
```{r}
library(tidyverse)
```

Table 2.1
```{r}
Y0 <- c(10,15,20,20,10,15,15)
Y1 <- c(15,15,30,15,20,15,30)
df<-as.data.frame(cbind(Y0,Y1))
```

Simulated treatment effects
```{r}
n <- nrow(df)

results <- data.frame(
  treated_case = integer(n),
  estimated_treatment_effect = numeric(n)
)

for(i in 1:7){
  treated_mean <- mean(df$Y1[i])
  
  control_mean <- mean(df$Y0[-i])
  
  results$treated_case[i] <- i
  results$estimated_treatment_effect[i] <- treated_mean - control_mean
}

results

mean(results$estimated_treatment_effect) # Answer: 5
```

#### (c) Show that the standard deviation  of the seven estimates is identical to the standard error implied by equation (3.4).

Standard deviation is 7.30 and the standard error, as calculated by equation 3.4 also results in 7.30. (See code below for calculations)

```{r}
# standard deviation of seven estimates
sd(results$estimated_treatment_effect)

# equation 3.4
n <- 7

m <- 1

var_Y0 <- var(df$Y0)

var_Y1 <- var(df$Y1)

cov_Y0Y1 <- cov(df$Y0, df$Y1)

sqrt((1/(n-1))*((m*var_Y0)/(n-m)+(((n-m)*var_Y1)/(m))+2*cov_Y0Y1))
```


#### (d) Referring to equation (3.4), explain why this experimental design has more sampling variability than the design in which two villages out of seven are assigned to treatment.

Since $Var(Y_i(0))$ is smaller (16.70) than $Var(Y_i(1))$ (50), increasing the number of selected villages m from 1 to 2, would increase the weight of the variance of $Y_i(0)$ in the equation from $\dfrac{1}{6}$ to $\dfrac{2}{5}$ (with formula of this weight being $\dfrac{m}{N-m}$) and decrease the weight of the variance of $Y_i(1)$ from $\dfrac{6}{1}$ to $\dfrac{5}{2}$. Since the variance of $Y_i(0)$ is smaller, this would result in a decrease in the sampling variability.

#### (e) Explain why, in this example, a design in which one of seven observations is assigned to treatment has more sampling variability than a design in which six villages out of seven are assigned to treatment.

See answer to question d, since the sample logic applies here. 