---
title: "Problem Set 3"
subtitle: "Causal Inference and Experiments in the Social Sciences"
author: "Thales Souza Lima"
date: today
format:
  pdf:
    documentclass: article
    geometry:
      - margin=1in
    number-sections: false
    keep-tex: false
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

# Exercise 3.5

c\)

```{r}
# Potential outcomes
y0 <- c(10, 15, 20, 20, 10, 15, 15)
y1 <- c(15, 15, 30, 15, 20, 15, 30)

# Number of villages and number assigned to treatment
N <- length(y0)
m <- 1

# Difference-in-means estimate for each possible randomization
ate_estimates <- sapply(seq_len(N), function(treated_village) {
  
  treated_mean <- y1[treated_village]
  control_mean <- mean(y0[-treated_village])
  
  treated_mean - control_mean
})

ate_estimates
mean(ate_estimates)

# Standard deviation of the complete randomization distribution.
# Because these seven assignments constitute the entire distribution,
# the denominator is N, rather than N - 1.
sd_randomization <- sqrt(
  mean((ate_estimates - mean(ate_estimates))^2)
)

sd_randomization

# Finite-population variance and covariance:
# Equation (3.4) defines these quantities using denominator N.
var_y0 <- mean((y0 - mean(y0))^2)
var_y1 <- mean((y1 - mean(y1))^2)

cov_y0_y1 <- mean(
  (y0 - mean(y0)) * (y1 - mean(y1))
)

# Standard error from Equation (3.4)
se_equation_3_4 <- sqrt(
  (1 / (N - 1)) *
    (
      (m * var_y0) / (N - m) +
      ((N - m) * var_y1) / m +
      2 * cov_y0_y1
    )
)

se_equation_3_4

# Verify that the two quantities are identical
all.equal(sd_randomization, se_equation_3_4)
```

d\) Because we are assigning only one village to treatment, the treatment mean is based on a single value. Taking from the equation, increasing N would decrease the treatment variance component.

e\) Assuming that the variance of the outcomes were the same, assigning 6 villages to treatment would have the same problem as assigning just 1. You are weighting one side more. This exercise has a different implication because the variance of the control group is smaller (as usual), therefore "adding more weights" to the group with the smaller variance would yield a smaller error.
